Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-21 papers/paper/ms.tex
update ms

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is measured
  from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  (harboring \nplwgyroagenograzing\ known planets) for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.  We find that the scarcity of
  sub-gigayear systems can be attributed to the star formation rate in
  the Kepler field dropping by a factor of \ratiosfr$\pm$\uncratiosfr\
  over the past 3\,Gyr.  We observe this trend both in known planet
  hosts and in the parent stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star formation history
  has been previously reported, and its confirmation helps clarify the
  age distribution of the known transiting exoplanets.
  %
\end{abstract}

@@

While stellar ensembles provide the gold standard for astronomical ages,
this ``cluster-first'' approach has a major limitation: most 
stars are in the field.  Only $\lesssim$1\% of stars within
$\approx$500\,pc have been associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
The implication for planetary ages can appreciated by querying the NASA Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Requiring $t$$<$1\,Gyr at 2$\sigma$ precision gives
$\approx$50 such planets at the time of writing.  Most of these young
planets are in clusters, and were found by Kepler, K2, or TESS.
However, these surveys have cumulatively discovered a total of $\approx$5{,}000 planets.
Assuming a constant star formation history, we would expect an
order of magnitude more sub-gigayear transiting planets.

@@

We calculated gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}, which is a method designed to address the fact
that stars with the same mass and same rotation period can have a wide
range of ages \citep[e.g.][]{Gallet_Bouvier_2015}.  The gyrochrone age
posterior should therefore incorporate the intrinsic population-level
scatter into its statistical precision.  Figure~\ref{fig:models}
highlights this problem, particularly in regions where the 0.1--1\,Gyr
stars overlap.

To estimate the stellar spin-down rate as a function of time and
stellar temperature, \texttt{gyro-interp} uses measured rotation
periods and effective temperatures from reference open clusters, and
interpolates between them using cubic Hermite polynomials.  We
calculated the probability of the rotation-based age, $t_{\rm gyro}$,
given the observed periods and temperatures (and their uncertainties)
by integrating Equation~1 of \citet{Bouma_2023}.  This procedure
marginalizes over the astrophysical scatter that is observed in the
open cluster sequences.  We adopted a linear age grid spanning
0--5\,Gyr with 500 grid points, and interpolated the resulting
posteriors to calculate any summary statistics.  The oldest cluster
for which \texttt{gyro-interp} is calibrated is the 4\,Gyr M67 cluster
\citep[][]{Dungee_2022,Gruner_2023}.  For \deleted{Kepler }stars with rotation
periods above the M67 sequence, our resulting age constraints are\added{ therefore }
quoted as lower limits.
\added{Finally, we opted to impose a $t_{\rm gyro}$ uncertainty floor of 10\%. 
This seemed warranted given the expected amplitude of differential rotation \citep[e.g.][]{Epstein_2014}, and
given the accuracy with which the absolute cluster age scale is currently known \citep[e.g.][]{Dahm_2015}. }

@@

Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
confirmed planets for which a sub-gigayear age is reported by at least
one age indicator.  All but two systems are either evolved stars or
else unresolved binaries, and are automatically flagged as such.  The
more interesting of the two is Kepler-786, an early K dwarf with
$t_{\rm Li}$=\kepseveneightsix\,Myr, but with a $\approx$33\,day
rotation period that implies $t_{\rm
gyro}$=\kepseveneightsixgyro\,Myr.  Other age indicators in the
spectrum similarly suggest that the star is not young.  It 
therefore appears to be anomalously lithium-rich.

@@

\added{A separate plausible origin for rotational outliers is that they come from
erroneously reported rotation periods.  While
the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
the referee correctly pointed out that this only a comparative test of methods,
rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
longer rotation periods.  In this scenario, our analysis would overestimate the
rate by which the SFR in the Kepler field has decreased, by 7-14\%. }

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet-hosts show a
    demographic cliff.  There are twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago.  This result from
    rotation-based ages broadly agrees with recent reports of a
    declining star-formation rate from CMD fitting and white-dwarf
    chronology, though with the advantage of better sensitivity for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    implies that we should instead expect $\approx$250.
  %
  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

>>>>

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is measured
  from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  (harboring \nplwgyroagenograzing\ known planets) for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.  We find that the scarcity of
  sub-gigayear systems can be attributed to the star formation rate in
  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \ratiosfr$\pm$\uncratiosfr\
  over the past 3\,Gyr.  We observe this trend both in known planet
  hosts and in the parent stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star formation history
  has been previously reported, and its confirmation helps clarify the
  age distribution of the known transiting exoplanets.
  %
\end{abstract}

@@

While stellar ensembles provide the gold standard for astronomical ages,
this ``cluster-first'' approach has a major limitation: most 
stars are in the field.  Only $\lesssim$1\% of stars within
$\approx$500\,pc have been associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
The implication for known planetary ages can appreciated by querying the NASA Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Requiring $t$$<$1\,Gyr at 2$\sigma$ precision gives
$\approx$50 such planets at the time of writing.  Most of these young
planets are in clusters, and were found by Kepler, K2, or TESS.
However, these surveys have cumulatively discovered \deleted{a total of }$\approx$5{,}000 planets.
Assuming a constant star formation history, we would expect an
order of magnitude more sub-gigayear transiting planets\added{ than the 50 currently known}.

@@

We calculated gyrochrone ages using \texttt{gyro-interp}
\citep[\added{\texttt{v0.6};}][]{Bouma_2023}, which is a method designed to address the fact
that stars with the same mass and same rotation period can have a wide
range of ages \citep[e.g.][]{Gallet_Bouvier_2015}.  The gyrochrone age
posterior should therefore incorporate the intrinsic population-level
scatter into its statistical precision.  Figure~\ref{fig:models}
highlights this problem, particularly in regions where the 0.1--1\,Gyr
stars overlap.

To estimate the stellar spin-down rate as a function of time and
stellar temperature, \texttt{gyro-interp} uses measured rotation
periods and effective temperatures from reference open clusters, and
interpolates between them using cubic Hermite polynomials.  We
calculated the probability of the rotation-based age, $t_{\rm gyro}$,
given the observed periods and temperatures (and their uncertainties)
by integrating Equation~1 of \citet{Bouma_2023}.  This procedure
marginalizes over the astrophysical scatter that is observed in the
open cluster sequences.  We adopted a linear age grid spanning
0--5\,Gyr with 500 grid points, and interpolated the resulting
posteriors to calculate any summary statistics.  The oldest cluster
for which \texttt{gyro-interp} is calibrated\added{ in \texttt{v0.6}} is the 4\,Gyr M67 cluster
\citep[][]{Dungee_2022,Gruner_2023}.  For \deleted{Kepler }stars with rotation
periods above the M67 sequence, our resulting age constraints are\added{ therefore} quoted as lower limits.

\added{Finally, we opted to modify \texttt{gyro-interp} in \texttt{v0.6} to
incorporate an age-dependent population-level Gaussian scatter, as detailed in the change log at \url{gyro-interp.readthedocs.io}.
While age precisions at $\leq$1\,Gyr are unaffected, this update
accounts for the observed intrinsic scatter in
Ruprecht-147 and M67 being larger than in Praesepe or NGC-6811,
potentially due to differential rotation \citep[e.g.,][]{Epstein_2014}.  
For $\approx$2-4\,Gyr old stars with rotation periods measured at $\approx$5\% precision, 
this change yields a $\lesssim$50\% increase in age uncertainties.}

@@

Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
confirmed planets for which a sub-gigayear age is reported by at least
one age indicator.  All but two systems are either evolved stars or
else unresolved binaries, and are automatically flagged as such.  The
more interesting of the two is Kepler-786, an early K dwarf with
$t_{\rm Li}$=\kepseveneightsix\,Myr, but with a $\approx$33\,day
rotation period that implies $t_{\rm
gyro}$$\approx$4.4,Gyr.  Other age indicators in the
spectrum similarly suggest that the star is not young.  It 
therefore appears to be anomalously lithium-rich.

@@

\added{A separate plausible origin for rotational outliers is that they come from
erroneously reported rotation periods.  While
the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
the referee correctly pointed out that this is only a comparative test of methods,
rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
longer rotation periods.  In this scenario, our analysis would overestimate the
degree by which the SFR in the Kepler field has decreased, by 7-14\%. }

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet-hosts show a
    demographic cliff.  There are twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago.  This result from
    rotation-based ages broadly agrees with recent reports of a
    declining star-formation rate from CMD fitting and white-dwarf
    chronology, though with the advantage of better sensitivity for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we should instead expect $\approx$250.
  %
  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}
